---
title: "Stat 423 Final Project"
author: "Josie Czeskleba"
date: "2025-02-27"
output: pdf_document
---

# Setup
```{r setup}
library(tidyverse)
library(corrplot)
library(bestNormalize)
```



# Read and Preprocess Data
```{r read data}

sleep <- read.csv("sleep_cycle_productivity.csv")

sleep$Gender <- as.factor(sleep$Gender)

rating_vars <- c("Sleep.Quality", "Productivity.Score", "Mood.Score", "Stress.Level")

for (var in rating_vars) {
  new_var <- paste0(var, "Cat")
  sleep[[new_var]] <- cut(sleep[[var]], breaks = c(0, 3, 7, 10),
                       labels = c("Low", "Medium", "High"), right = TRUE)
  sleep[[new_var]] <- as.factor(sleep[[new_var]])
}


head(sleep)
```

# Exploratory Data Analysis
```{r eda}
hist(sleep$Total.Sleep.Hours, main = "Histogram of Total Sleep Hours")
plot(sleep$Caffeine.Intake..mg., sleep$Total.Sleep.Hours, main = "Total Sleep Hours vs Caffeine Intake")
plot(sleep$Screen.Time.Before.Bed..mins., sleep$Total.Sleep.Hours, main = "Total Sleep Hours vs Screen Time")
plot(sleep$Work.Hours..hrs.day., sleep$Total.Sleep.Hours, main = "Total Sleep Hours vs Total Work Hours")
```

# Model Building

## Non-Transformed Simple Models
```{r simple models}
model1 <- lm(Total.Sleep.Hours ~ (Age + Gender + Exercise..mins.day. + Caffeine.Intake..mg. +
               Screen.Time.Before.Bed..mins. + Work.Hours..hrs.day. + Sleep.QualityCat + 
               Productivity.ScoreCat + Mood.ScoreCat + Stress.LevelCat)^2, data = sleep)
model2 <- lm(Work.Hours..hrs.day. ~ (Age + Gender + Exercise..mins.day. + Caffeine.Intake..mg. +
               Screen.Time.Before.Bed..mins. + Total.Sleep.Hours + Sleep.QualityCat + Productivity.ScoreCat +
               Mood.ScoreCat + Stress.LevelCat)^2, data = sleep)
summary(model1)
summary(model2)


# Plot Residuals 

plot(fitted(model1), residuals(model1), 
     main = "Model 1Residuals vs Fitted", 
     xlab = "Fitted Values", 
     ylab = "Residuals") 
abline(h = 0, col = 'red')
qqnorm(residuals(model1), main = "QQplot for Model 1")
qqline(residuals(model1), col = 'red')

plot(fitted(model2), residuals(model2), 
     main = "Model 2 Residuals vs Fitted", 
     xlab = "Fitted Values", 
     ylab = "Residuals") 
abline(h = 0, col = 'red')
qqnorm(residuals(model2), main = "QQplot for Model 2")
qqline(residuals(model2), col = 'red')



```
There is a strong S shape visible in both QQ Plots. This implies that the data deviates significantly from the assumed normal distributed. 

## Normalizing Response Variables 
Since it is unclear what relationship is present in our data we will use the Box Cox method to select the best transformation. 

```{r boxcox + quantile normalization}
bc_sleep <- boxcox(sleep$Total.Sleep.Hours)
sleep$normalized_sleep <- predict(bc_sleep)

bn_sleep <- bestNormalize(sleep$Total.Sleep.Hours)
sleep$bn_sleep <- predict(bn_sleep)

bc_work_hours <- boxcox(sleep$Work.Hours..hrs.day.)
sleep$normalized_work_hours <- predict(bc_work_hours)

bn_work_hours <- bestNormalize(sleep$Total.Sleep.Hours)
sleep$bn_work_hours <- predict(bn_work_hours)

par(mfrow = c(3,3))
hist(sleep$Total.Sleep.Hours, main = "Total Sleep Hours")
hist(sleep$normalized_sleep, main = "Box-Cox Sleep Hours")
hist(sleep$bn_sleep, main = "QN Sleep Hours")
hist(sleep$Work.Hours..hrs.day., main = "Work Hours per Day")
hist(sleep$normalized_work_hours, main = "Box-Cox Work Hours")
hist(sleep$bn_work_hours, main = "QN Work Hours")

```


## Transformed Two-Way Interacition Models
```{r boxcox model}
model3 <- lm(normalized_sleep ~ (Age + Gender + Exercise..mins.day. + Caffeine.Intake..mg. +
               Screen.Time.Before.Bed..mins. + Work.Hours..hrs.day. + Sleep.QualityCat + 
               Productivity.ScoreCat + Mood.ScoreCat + Stress.LevelCat)^2, data = sleep)
model4 <- lm(normalized_work_hours ~ (Age + Gender + Exercise..mins.day. +
                                     Caffeine.Intake..mg. + Screen.Time.Before.Bed..mins. +
                                     Total.Sleep.Hours + Sleep.QualityCat + Productivity.ScoreCat +
                                     Mood.ScoreCat + Stress.LevelCat)^2, data = sleep)

summary(model3)
summary(model4)

par(mfrow = c(2,2))

plot(fitted(model3), residuals(model3), 
     main = " Model 3 Residuals vs Fitted", 
     xlab = "Fitted Values", 
     ylab = "Residuals") 
abline(h = 0, col = 'red')
qqnorm(residuals(model3), main = "QQplot for Model 3")
qqline(residuals(model3), col = 'red')

plot(fitted(model4), residuals(model4), 
     main = "Model 4 Residuals vs Fitted", 
     xlab = "Fitted Values", 
     ylab = "Residuals") 
abline(h = 0, col = 'red')
qqnorm(residuals(model4), main = "QQplot for Model 4")
qqline(residuals(model4), col = 'red')
```

```{r quantile normalized model}

model5 <- lm(bn_sleep ~ (Age + Gender + Exercise..mins.day. + Caffeine.Intake..mg. +
               Screen.Time.Before.Bed..mins. + Work.Hours..hrs.day. + Sleep.QualityCat + 
               Productivity.ScoreCat + Mood.ScoreCat + Stress.LevelCat)^2, data = sleep)
model6 <- lm(bn_work_hours ~ (Age + Gender + Exercise..mins.day. + Sleep.Start.Time +
                                     Caffeine.Intake..mg. + Screen.Time.Before.Bed..mins. +
                                     Total.Sleep.Hours + Sleep.QualityCat + Productivity.ScoreCat +
                                     Mood.ScoreCat + Stress.LevelCat)^2, data = sleep)

summary(model5)
summary(model6)

par(mfrow = c(2,2))

plot(fitted(model5), residuals(model5), 
     main = " Model 5 Residuals vs Fitted", 
     xlab = "Fitted Values", 
     ylab = "Residuals") 
abline(h = 0, col = 'red')
qqnorm(residuals(model5), main = "QQplot for Model 5")
qqline(residuals(model5), col = 'red')

plot(fitted(model6), residuals(model6), 
     main = "Model 6 Residuals vs Fitted", 
     xlab = "Fitted Values", 
     ylab = "Residuals") 
abline(h = 0, col = 'red')
qqnorm(residuals(model6), main = "QQplot for Model 6")
qqline(residuals(model6), col = 'red')
```





